A Formal Treatment of Bidirectional Typing

### 2024-01-18 tex/ESOP24/S6-example.tex
Sprinkled the paper with commas; miscellaneous revisions

\subsection{Spine application}\label{subsec:spine}
A spine application $t\;u_1\;\ldots\;u_n$ is a form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
This arrangement allows direct access to the head term, making it practical in various applications, and has been used by \Agda's core language. %employs this form of application, as does its reflected syntax used for metaprogramming.

@@

\subsection{Computational calculi}\label{subsec:PCF}
Implementing a stand-alone type synthesiser for a simply typed language is typically a straightforward task.
However, the code size increases proportionally to the number of type constructs and of arguments associated with each term construct.
When dealing with a fixed number $n$ of type constructs, for each synthesising construct, there are two cases for a checking argument but $n + 1$ cases for each synthesising argument---the successful synthesis of the expected type, an instance where it fails, or $n-1$ cases where the expected type does not match; similarly for a checking construct---making the task tedious.
Thus, having a generator is helpful and can significantly reduce the effort for implementation.

@@

On the other hand, similar to a parser generator, a type-synthesiser generator only needs a specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
Such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
\[
  \arity({\mathsf{nat}}) = 0\qquad
  \arity({\mathsf{T}}) = 1\qquad
  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2.
\]
Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are given as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
Mode-correctness can be derived by \cref{lem:decidability-mode-correctness} and its type synthesiser by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$ directly.
For a demonstration in \Agda, see \cref{sec:formalisation}.
\else
For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with naturals, sums, products, and general recursion.
The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, whereas there are nearly 100 possible cases to consider in bidirectional type synthesis.
However, it has far fewer constructs than a realistic programming language would have.

>>>>

\subsection{Spine application}\label{subsec:spine}
A spine application $t\;u_1\;\ldots\;u_n$ is a form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1$, $u_2$, \ldots, $u_n$.
This arrangement allows direct access to the head term, making it practical in various applications, and has been used by \Agda's core language. %employs this form of application, as does its reflected syntax used for metaprogramming.

@@

\subsection{Computational calculi}\label{subsec:PCF}
Implementing a stand-alone type synthesiser for a simply typed language is typically a straightforward task.
However, the code size increases proportionally to the number of type constructs and of arguments associated with each term construct.
For example, when dealing with a fixed number $n$ of type constructs, for each synthesising construct there are two cases for a checking argument but $n + 1$ cases for each synthesising argument: the successful synthesis of the expected type, an instance where it fails, or $n-1$ cases where the expected type does not match.
Thus, having a generator is helpful and can significantly reduce the effort for implementation.

@@

On the other hand, similar to a parser generator, a type-synthesiser generator only needs a specification $(\Sigma, \Omega)$ from the user to produce a corresponding synthesiser. 
Such a specification can be derived by extending $\Sigma_{\bto}$ with additional type constructs with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
\[
  \arity({\mathsf{nat}}) = 0\qquad
  \arity({\mathsf{T}}) = 1\qquad
  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2.
\]
Types $\mathtt{nat}$, $A \times B$, $A + B$, and $T(A)$ are given as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
The signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations listed in \cref{tab:computational-calculus}.
Mode-correctness can be derived by \cref{lem:decidability-mode-correctness} and its type synthesiser by \cref{cor:trichotomy} with the specification~$(\Sigma, \Omega)$ directly.
For a demonstration in \Agda, see \cref{sec:formalisation}.
\else
For illustrative purposes, consider a computational calculus~\cite{Moggi1989} with naturals, sums, products, and general recursion.
The extended language has `only' 15 constructs, including pairing, projections, injections, and so on, and this number of constructs is still far fewer than what a realistic programming language would have.
Even for this small calculus, there are already nearly 100 possible cases to consider in bidirectional type synthesis.

### 2024-01-18 tex/ESOP24/S8-future.tex
Sprinkled the paper with commas; miscellaneous revisions

The idea of presenting logics universally at least date back to universal algebra and model theory where structures are studied for certain notions of arities and signatures.
In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which has been used to prove a general confluence theorem.
Many general definitions and frameworks for defining logics and type theories have been proposed and can be classified into two groups by where signatures reside: the meta level or the object level of a meta-language:
\begin{enumerate}
  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
\item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory) where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
\end{enumerate}

The \LF approach is naturally expressive but each extension requires a different framework and a different implementation to check formal \LF proofs.
Formalising \LF and its variants is at least as hard as formalising a dependent type theory, and they are mostly implemented separately from their theory and unverified.

For the second group, theories developed in set theory can often be restated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
Such examples include frameworks developed by \citet{Allais2021,Ahrens2022,Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is limited to simply typed theories.
Our work belongs to the second group, as we aim for a formalism in a type theory to minimise the gap between theory and implementation.

@@

\ifarxiv
However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}:
\[
  \inferrule{\Gamma \vdash t \syn \forall \alpha. A}{\Gamma \vdash t \syn A[B/\alpha]}
\]
\else
However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}
\fi
which is not syntax-directed.
By translating the rule to subtyping, we have the \emph{instantiation problem} which amounts to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
A theory that accommodates various solutions to the problem is left as future work.

@@

Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
They propose a notion of \LF signature with modes, a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of bidirectional type synthesis.
The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and strongly normalising.
The last assumption, however, is challenging to establish even for specific languages and lacks a general understanding.

@@

To explore general settings, an extreme choice is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.

Yet, we could first focus on simpler cases that the ordinary typing part is still syntax-directed, but each typing rule is refined to multiple bidirectional variants, including different orders of its premises.
In such cases, the mode decorator would need to find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.
Completeness could still take the simple form presented in this paper too.

The next step is to consider non-syntax-directed systems where each typing rule has multiple bidirectional variants.
For this setting, the mode decorator and the bidirectional type synthesiser will have to backtrack.
It is still possible to treat soundness as the separation of mode and type information, but completeness will pose a problem.
For every raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
To retain the mode decoration, one could make the mode decorator produce all possible mode derivations, and refine completeness to say that any typing derivation can be combined with one of mode derivations into a bidirectional typing derivation.
%It should be possible to fix this problem while retaining the mode decoration phase.

>>>>

The idea of presenting logics universally at least date back to universal algebra and model theory, where structures are studied for certain notions of arities and signatures.
In programming language theory, \varcitet{Aczel1978}{'s binding signature} is an example which has been used to prove a general confluence theorem.
Many general definitions and frameworks for defining logics and type theories have been proposed and can be classified into two groups by where signatures reside: the meta level or the object level of a meta-language:
\begin{enumerate}
  \item \varcitet{Harper1993a}{'s logical framework \LF} and its family of variants~\cite{Harper2007,Assaf2016,Felicissimo2023} are \emph{extensions} of Martin-L\"of type theory, where signatures are on the \emph{meta level} and naturally capable of specifying dependent type theories;
\item general dependent type theories~\cite{Bauer2020,Haselwarter2021,Bauer2022a,Uemura2021}, categorical semantics~\cite{Fiore1999,Tanaka2006,Tanaka2006a,Fiore2010,Hamana2011,Fiore2013,Arkor2020,Fiore2022} (which includes the syntactic model as a special case), and frameworks for substructural systems~\cite{Tanaka2006,Tanaka2006a,Wood2022} are developed \emph{within} a meta-theory (set theory or type theory), where signatures are on the \emph{object level} and their expressiveness varies depending on their target languages.
\end{enumerate}

The \LF approach is naturally expressive, but each extension requires a different framework and a different implementation to check formal \LF proofs.
Formalising \LF and its variants is at least as hard as formalising a dependent type theory, and they are mostly implemented separately from their theory and unverified.

For the second group, theories developed in set theory can often be restated in type theory and thus manageable for formalisation in a type-theoretic proof assistant. 
Such examples include frameworks developed by \citet{Ahrens2022} in \Coq and by \citet{Allais2021} and \citet{Fiore2022} in \Agda, which are relatively correct to their meta-language, although their expressiveness is currently limited to simply typed theories.

Our work belongs to the second group, as we aim for a formalism in a type theory to minimise the gap between theory and implementation.

@@

\ifarxiv
However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}:
\[
  \inferrule{\Gamma \vdash t \syn \forall \alpha. A}{\Gamma \vdash t \syn A[B/\alpha]}
\]
\else
However, explicit type application in \SystemF and \SystemFsub is impractical but its implicit version results in a \emph{stationary rule}~\cite{Leivant1986}
\fi
which is not syntax-directed.
By translating the rule to subtyping, we have the \emph{instantiation problem} that requires guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
A theory that accommodates various solutions to the problem is left as future work.

@@

Extending the logical framework \Dedukti~\cite{Assaf2016}, \varcitet{Felicissimo2023}{'s recent \textsf{CompLF}} is capable of defining dependent type theories with computational rules and enables generic bidirectional type synthesis.
They propose a notion of \LF signature with modes, a definition of mode-correctness (called \emph{well-typed} op.\ cit.), and an algorithm of bidirectional type synthesis.
The algorithm is decidable, sound and complete with respect to mode-decorated terms, provided that the specified language is mode-correct, and that the set of computational rules is type-preserving, confluent, and strongly normalising.
The last assumption, however, is challenging to establish even for specific languages, and lacks a general treatment.

@@

%To explore general settings, an extreme choice is to independently specify an ordinary and a bidirectional type system over the same raw terms and then investigate relationships between the two systems.

To relax the assumption of syntax-directedness, we could start from a simpler case where the ordinary typing part is still syntax-directed, but each typing rule is refined to multiple bidirectional variants, including different orders of its premises.
In such cases, the mode decorator would need to backtrack and find all mode derivations, but the type synthesiser should still work in a syntax-directed manner on each mode derivation.
Completeness could still take the simple form presented in this paper too.

Next, we could consider systems where each construct can have multiple typing rules, which can have multiple bidirectional variants.
In this setting, the bidirectional type synthesiser will also need to backtrack.
It is still possible to treat soundness as the separation of mode and type information, but completeness will pose a problem:
for every raw term, a mode derivation chooses a mode assignment while a typing derivation chooses a typing rule, but there may not be a bidirectional typing rule for this particular combination.
A solution might be refining completeness to say that any typing derivation can be combined with one of the possible mode derivations into a bidirectional typing derivation.
%It should be possible to fix this problem while retaining the mode decoration phase.
